Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-10-17 boettiger.tex
PNAS submission ready

                         \newcommand{\ud}{\mathrm{d}}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% Don't type in anything in the following section:
                         %%%%%%%%%%%%
                         %% For PNAS Only:
                         \contributor{Submitted to Proceedings
                         of the National Academy of Sciences of the United States of America}
                         \url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
                         \copyrightyear{2008}
                         \issuedate{Issue Date}
                         \volume{Volume}
                         \issuenumber{Issue Number}
                         %%%%%%%%%%%%

@@

Catastrophic regime shifts in complex natural systems may be averted through advanced detection. 
Recent work has provided a proof-of-principle that many systems approaching a catastrophic transition look different 
through the lens of early warning indicators such as rising variance or increased return times.  
Without the benefit of hindsight of a collapse, 
applications of these approaches must quantify how reliable different indicators are in avoiding false alarms, 
and how sensitive they are to missing subtle warning signs.  
We propose an approach which quantifies this trade-off between reliability and sensitivity, 
allows comparisons between different indicators, 
and estimates what data are necessary to achieve a desired error rate.  
We also introduce a likelihood-based indicator that is more sensitive than previously proposed summary statistics.

@@

\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011} , and semi-arid ecosystems~\cite{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009}.
but in cases where good models are not available general approaches are needed~\cite{Scheffer2009}⁠.
We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system~\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}}.  
This meets the assumptions and thus provides a ``best-case'' test of detection approaches,
yet we will demonstrate the resulting uncertainty can remain very high.
We describe a novel statistical approach for handling this uncertainty that is appropriate for the goal of informing management.

Though previous research has identified patterns such as increasing variance or autocorrelation in time-series approaching a critical transition,
there has been no method for forecasting the probability of a collapse from a natural system.
As real-world applications of early warning signals will not be able to point to the same differences between controlled and
manipulated systems that have been demonstrated in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
and fewer still that determine the distribution of that quantity in the event that the system is not approaching a transition --
the reliability or false alarm rate~\cite{Dakos2008}⁠. 
None have considered the distribution of such a statistic in the event that a system is approaching a transition -- 
the sensitivity to avoid missed detection.  
We need to quantify such uncertainty/error rates to compare between possible methods~\cite{Contamin2009}
and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.

Detection of impending transitions is not just a scientific question and the goal should not solely be to detect with a 5\% chance of a false positive,
and a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
with which it will fail to provide any advanced warning. 
Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
In this paper, we
(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of an insensitive method 
(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.

@@

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
% so this is not a reasonable statistical characterization
visual or heuristic measure of increase\footnote{That is, do not provide $p$-values or a statistic for which they could be generated.} in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
In principle, other quantifications of increasing variance, autocorrelation, etc.,
may perform better than $\tau$ and could be analyzed just as well in this framework. 
While we use this established metric of increase, we also introduce a novel warning indicator
that is not based on the gradual increase of a summary statistic but on likelihood comparisons.

Having identified an indicator statistic,
we must be able to generate its distribution under both the scenario of a stable system and 
one approaching a transition before we can compute the ROC curve.
We fit general stochastic models representing common bifurcations
(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010} )
for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
By estimating the statistic (e.g. $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of $\tau$,
which provides a more direct and more powerful test than detecting an increase in summary statistics.
As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.

@@

By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
Other literature has already begun to address the additional challenges posed when the underlying 
dynamics do not correspond to these models~\cite{Hastings2010}.
Our results illustrate that even in this best-case scenario, 
reliable identification of warning signals from summary statistics can be difficult.  
The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
or even the distribution of that statistic under the null model of a stable system, 
but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
By estimating the ROC curve for a given set of data, 
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
We can develop indicators that get the most from this data.

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model-adequacy.  
This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
and systems that violate these assumptions can exhibit sudden transitions without any warning,
such as chaotic systems or those with non-smooth potential functions~\cite{Hastings2010}.
Sudden transitions have also been observed in ecological models in which 
the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
As the bifurcation results from the principle eigenvalue passing through zero, 
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.  
The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
(introduced by the Poisson process demographic events), 
demonstrates the accuracy under violation of these assumptions.

@@

@@

Acknowledgements
S. Schreiber, M. Holyoak, M. Baskett, and A. Perkins provided comments on the manuscript.  This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.

@@

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig3.pdf}
     \label{fig3}
     \caption{ROC curves for different early warning indicators on simulated and empirical data.  The area under the curve, inset, indicates how the method compares to a random guess (0.5).  The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data. The simulated dataset has 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121. See corresponding distributions, Fig. S1, and additional examples, Fig. S2.}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig4.pdf}
     \label{fig4}
     \caption{ROC curves (from Fig. 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.  Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.} 
  \end{center}
 \end{figure*}

>>>>

                         \newcommand{\ud}{\mathrm{d}}
                         \renewcommand{\thefootnote}{\fnsymbol{footnote}}
                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% Don't type in anything in the following section:
                         %%%%%%%%%%%%
                         %% For PNAS Only:
                         \contributor{Submitted to Proceedings
                         of the National Academy of Sciences of the United States of America}
                         \url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
                         \copyrightyear{2008}
                         \issuedate{Issue Date}
                         \volume{Volume}
                         \issuenumber{Issue Number}
                         %%%%%%%%%%%%

@@

Catastrophic regime shifts in complex natural systems may be averted through advanced detection. 
Recent work has provided a proof-of-principle that many systems approaching a catastrophic transition may be identified 
through the lens of early warning indicators such as rising variance or increased return times.  
Without the benefit of hindsight of a collapse, 
applications of these approaches must quantify how reliable different indicators are in avoiding false alarms, 
and how sensitive they are to missing subtle warning signs.  
We propose an approach which quantifies this trade-off between reliability and sensitivity, 
allows comparisons between different indicators, 
and estimates what data are necessary to achieve a desired error rate. 
We show these error rates can be quite severe for common indicators even under favorable assumptions, 
and illustrate how a likelihood-based indicator can improve this performance.

@@

\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}  
Making this assumption provides a ``best-case'' test of detection approaches;
we will demonstrate the resulting uncertainty can be very high.
We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.

Though previous research has identified patterns such as increasing variance or autocorrelation
in time-series approaching a critical transition,
there has been no method for forecasting the \emph{probability} of a collapse from a natural system.
Real-world applications of early warning signals will not be able to point to such differences between controlled and
manipulated systems that have been utilized in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
and fewer still that statistically estimate the false alarm rate~\cite{Dakos2008} or 
the probability of missed detection⁠. 
We need to quantify such uncertainty and error rates to compare between possible methods~\cite{Contamin2009}
and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.

Detection of impending transitions is not just a scientific question 
and the goal should not solely be to detect with a 5\% chance of a false positive:
a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
with which it will fail to provide any advanced warning. 
Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
In this paper, we
(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of insensitive indicators 
(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.

@@

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
% so this is not a reasonable statistical characterization
visual or heuristic measure of increase\footnote{That is, do not provide $p$-values or a statistic for which they could be generated.} in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
In principle, other quantifications of increasing variance, autocorrelation, etc.,
may perform better than $\tau$ and could be analyzed just as well in this framework. 
While we use this established metric of increase, we also introduce a novel warning indicator
that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).

Having identified an indicator statistic,
we must be able to generate its distribution under both the scenario of a stable system and 
one approaching a transition before we can compute the ROC curve.
We fit general stochastic models representing common bifurcations
(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010})
for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
By estimating the statistic (e.g. $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of $\tau$,
which provides a more direct and more powerful test than detecting an increase in summary statistics.
As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.

@@

By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
Other literature has already begun to address the additional challenges posed when the underlying 
dynamics do not correspond to these models~\cite{Hastings2010}.
Our results illustrate that even in this best-case scenario, 
reliable identification of warning signals from summary statistics can be difficult.  
The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
or even the distribution of that statistic under the null model of a stable system, 
but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
By estimating the ROC curve for a given set of data, 
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop indicators that get the most from this data.

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model adequacy.  
This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
and systems that violate these assumptions can exhibit sudden transitions without any warning,
such as chaotic systems or those with non-smooth potential functions~\cite{Hastings2010}.
Sudden transitions have also been observed in ecological models in which 
the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
As the bifurcation results from the principle eigenvalue passing through zero, 
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.  
The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
(introduced by the Poisson process demographic events), 
demonstrates the accuracy under violation of these assumptions.

@@

S. Schreiber, M. Holyoak, M. Baskett, and A. Perkins provided comments on the manuscript.  This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.

@@

@@

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig3.pdf}
     \label{fig3}
     \caption{ROC curves for different early warning indicators on simulated and empirical data.  The area under the curve, inset, indicates how the method compares to a random guess (0.5).  The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data. The simulated data sets have 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121. See corresponding distributions, Fig. S1.}
  \end{center}
 \end{figure*}

 \begin{figure*}
   \begin{center}
     \includegraphics[width=\linewidth]{Fig4.pdf}
     \label{fig4}
     \caption{ROC curves (from Fig. 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.  Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective. A warning sighal is not detected in stable sytem simulation. Corresponding distributions shown in Fig S2-5.} 
  \end{center}
 \end{figure*}

### 2011-10-17 pnas_appendix.tex
PNAS submission ready

\journal{Proceedings of the National Academy of Sciences } 
\begin{document}
%\begin{frontmatter}
%  \title{  }
%  \author[cpb]{Carl Boettiger\corref{cor1}}
%  \author[alan]{Alan Hastings}
%  \ead{cboettig@ucdavis.edu}
%  \cortext[cor1]{Corresponding author.}
%  \address[cpb]{Center for Population Biology, University of California, Davis, United States}
%  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, United States}
%\end{frontmatter}
%

\section{Distributions of the indicator statistic}
The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.  The value of the statistic observed in the original data is indicated by a black triangle. Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.  Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.  The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS1.pdf}
  \end{center}
  \caption{Distribution of the indicator statistic}
  \label{fig:S1}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS2.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a stable system.}
  \label{fig:S2}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS3.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a system under deteriorating stability.}
  \label{fig:S3}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS4.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
  \label{fig:S4}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS5.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation data.}
  \label{fig:S5}
\end{figure}

\section{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

>>>>

\journal{Proceedings of the National Academy of Sciences } 
\begin{document}
%\begin{frontmatter}
%  \title{  }
%  \author[cpb]{Carl Boettiger\corref{cor1}}
%  \author[alan]{Alan Hastings}
%  \ead{cboettig@ucdavis.edu}
%  \cortext[cor1]{Corresponding author.}
%  \address[cpb]{Center for Population Biology, University of California, Davis, United States}
%  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, United States}
%\end{frontmatter}
%
\appendix
\renewcommand*\thefigure{S\arabic{figure}}
\renewcommand*\theequation{S\arabic{equation}}
\section{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

 \section{Distributions of the indicator statistic}
The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.  The value of the statistic observed in the original data is indicated by a black triangle. Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.  Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.  The large overlap between the distribution under the stable model (blue) and the deteriorating stability system (red) indicates how difficult it can be for such statistics to descriminate between these alternatives.  The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS1.pdf}
  \end{center}
  \caption{Distribution of the indicator statistics corresponding to the ROC curves in Fig. 3, main text. As in Fig. 1, the expected distribution of the statistic under the null hypothesis of a stable system is shown in blue, while the distribution under the estimated deteriorating stability system is shown in red.  The value of the test statistic ($\delta$ in Likelihood, top row, otherwise $\tau$), observed in the original timeseries is indicated by the black triangle. Distributions are produced from a Gaussian kernel density estimate from 500 replicates, as described in Methods.}
  \label{fig:S1}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS2.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a stable system, corresponding to the first column of Fig. 4, main text.  Distributions as described in caption of Fig~\ref{fig:S1}.}
  \label{fig:S2}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS3.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a system under deteriorating stability, corresponding to the second column of Fig. 4, main text.  Distributions as described in caption of Fig~\ref{fig:S2}.}
  \label{fig:S3}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS4.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data, corresponding to the third column of Fig. 4, main text.  Distributions as described in caption of Fig~\ref{fig:S1}.}
  \label{fig:S4}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS5.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation data, corresponding to the fourth column of Fig. 4, main text.  Distributions as described in caption of Fig~\ref{fig:S1}.}
  \label{fig:S5}
\end{figure}
